A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-25 manuscript.qmd
shorten missing data sentence

First, requirements of real-world studies are often more complex than the simplified designs assumed by many user-friendly software packages for sample size planning.
One frequent issue in applied data analysis is missing data [@littleStatisticalAnalysisMissing2014].
Data can be missing completely at random (e.g., because an electronic measurement device randomly failed for some technical reasons).
Alternatively, subjects might systematically drop out or produce missing data, but this dropout can be explained by some attributes also measured in the dataset (e.g., older subjects have a higher probability to refuse answering a question on income).
In a more complicated scenario, missing data in some variables is caused by the measured attribute itself (e.g., wealthy people are more likely to refuse to report their income).
Moreover, many experimental designs contain conditions in which values of the predictor variables are missing by design. This can make data analysis more complicated because predictors have to be coded in specific ways that prevent the estimated GLMM from becoming unidentified. Whether missing data has an effect on the sample size planning depends on our theoretical assumptions on how the missingness is caused.
However, it is often challenging to decide whether missing data can be safely ignored in the data analysis and sample size planning process based on a merely theoretical approach [@gomilaMissingDataExperiments2022].
Tailored simulation-based approaches offer the possibility to include the assumed process of how data become missing in the data simulation, thereby determining the required sample size based on simulated datasets that contain missing values [for example, see @lanePowerStrugglesEstimating2018]. As a byproduct, the simulated datasets can also be used to test whether the intended data analysis provides the expected (unbiased) results, despite the missing data.
Although GLMMs can handle a large variety of outcome variables, researchers are becoming increasingly aware that many datasets might profit from even more sophisticated models. Common examples are zero-inflated outcomes, censoring, and nonlinear predictor effects that can be modeled with the R packages *glmmTMB* [@brooksGlmmTMBBalancesSpeed2017] or *brms* [@burknerAdvancedBayesianMultilevel2018]. 
Tailored simulation-based approaches do not share the same limitations as the existing software solutions for power analysis that focus exclusively on GLMMs.
As long as there is a software package available to estimate the model of interest, it is always possible to perform tailored simulation-based sample size planning.

>>>>

First, requirements of real-world studies are often more complex than the simplified designs assumed by many user-friendly software packages for sample size planning.
One frequent issue in applied data analysis is missing data [@littleStatisticalAnalysisMissing2014].
Missings can be completely random (e.g., an electronic measurement device randomly failed for technical reasons), depend on some attributes also measured in the dataset (e.g., older subjects refuse answering a question on income), or be caused by the measured attribute itself (e.g., wealthy people refuse to report their income).
Moreover, many experimental designs contain conditions in which values of the predictor variables are missing by design. This can make data analysis more complicated because predictors have to be coded in specific ways that prevent the estimated GLMM from becoming unidentified. Whether missing data has an effect on the sample size planning depends on our theoretical assumptions on how the missingness is caused.
However, it is often challenging to decide whether missing data can be safely ignored in the data analysis and sample size planning process based on a merely theoretical approach [@gomilaMissingDataExperiments2022].
Tailored simulation-based approaches offer the possibility to include the assumed process of how data become missing in the data simulation, thereby determining the required sample size based on simulated datasets that contain missing values [for example, see @lanePowerStrugglesEstimating2018]. As a byproduct, the simulated datasets can also be used to test whether the intended data analysis provides the expected (unbiased) results, despite the missing data.
Although GLMMs can handle a large variety of outcome variables, researchers are becoming increasingly aware that many datasets might profit from even more sophisticated models. Common examples are zero-inflated outcomes, censoring, and nonlinear predictor effects that can be modeled with the R packages *glmmTMB* [@brooksGlmmTMBBalancesSpeed2017] or *brms* [@burknerAdvancedBayesianMultilevel2018]. 
Tailored simulation-based approaches do not share the same limitations as the existing software solutions for power analysis that focus exclusively on GLMMs.
As long as there is a software package available to estimate the model of interest, it is always possible to perform tailored simulation-based sample size planning.
